---
title: "Tutorial ch11-2 script"
author: "Ryan Huang"
date: "10/26/2019"
output:
  html_document:
    df_print: paged
    toc: true
    toc_depth: 4
    number_sections: true
    toc_float: true
---

```{r, results='hide'}
library(rethinking)
library(rstan)
library(tidyverse)
library(gridExtra)
library(skimr) # for n_unique func
library(DMwR) # for unscale func
compare = rethinking::compare
```

# CH11 Mixture and Monster

## Ordered categorical regression

### Ordered outcome
Sometimes we have outcome variables that is discrete but in which the values indicate different ordered levels. These ordered variables should be treated differently from discrete variables or continuous variables. It's like a multinomial classification problem with the constraint that the categories have a certain direction. Our problem is how to ensure that the model maps to the outcome in the correct direction.

The data we use here is from an experiment. People are asked how much permissible that action is taken in a trolley problem under different scenarios. We want to investigate how people's reaction change under 3 different setting: *action*, *intention* and *contact*.

```{r}
data("Trolley")
d_Trolley = Trolley
```

Let's look at the overall distribution of the response.
```{r}
d_Trolley %>% 
  ggplot() +
  geom_histogram(aes(response))
```


### Ordered predictor