---
title: "Tutorial1 Ch2"
author: "Ryan Huang"
date: "7/26/2019"
output:
  html_document:
    df_print: paged
    toc: true
    toc_depth: 4
    number_sections: true
    toc_float: true
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(rethinking)
library(rstan)
```

## Chapter 2
In ch2, the author illustrate the following idea:

`posterior = prior * likelihood`

by using an example of throwing a globe representing the Earth.


After nine tosses we get our `data`: W L W W W L W L W, (6W, 3L).

1. Goal: Estimate the proportion of water on Earth, `p`

2. How:
  + Model: data generating processes in our mind
  + Fit the model: different approach for estimating `p`
  + Evaluation
  
  
### Model

We assumed that the counts of **water** W and **land** L are distributed `binomially`, with probability `p` of **water** on each toss.

### Fit the model

There are different methods for us to estimate what's the most probable value of `p`:
- Method 1: Grid Approximation
- Method 2: Stan / MCMC

#### Grid Approximation:

1. define the grid
2. define the weight of every p (`prior`)
3. calculate the likelihood using `dbinom` (dbinom: the likelihood of having 6W from 9 draws given p)
4. posterior = prior * likelihood


```{r}
d = 
  tibble(
    p_grid = seq(from=0, to=1, length.out = 100),
    prior  = 1
  ) %>% 
  mutate(
    grid.likelihood = dbinom(6, size = 9, prob = p_grid)
  ) %>% 
  mutate(
    grid.unstd_post = prior * grid.likelihood
  ) %>% 
  mutate(
    grid.post = grid.unstd_post / sum(grid.unstd_post)
  )

d$grid.post %>% sum()


d %>% 
  ggplot() +
  geom_line(aes(p_grid, grid.post))
```













#### Stan / MCMC:
There is another approach called MCMC. MCMC will draw lots of samples to form the posterior, that is, we can calculate the count of samples to know the posterior.


1. **data**: 6W, 3L
2. **parameters**: `p`
3. **model**: W: binomial distribution; P: unifrom

```{r, results='hide'}
binom.model = "
data {
	int N; // number of sample
	int W; // number of suc
}
parameters {
	real<lower=0, upper=1> p;
}
model {
	// prior
	p ~ uniform(0, 1);

	//likelihood
	W ~ binomial(N, p);
}
"
binom.data = list(N = 9, W = 6)
binom.fit = stan(model_code = binom.model, 
                 data = binom.data,
                 chains = 2,
                 iter = 5000)
binom.fit
```

```{r}
binom.post = as.data.frame(binom.fit)
dens(binom.post$p)

a = binom.post$p %>% 
  hist(breaks = seq(from=0, to=1, length.out = 100), plot = F) %>% 
  .$counts %>% 
  c(.,0)

d = d %>% 
  mutate(
    stan.post = a / sum(a)
  )


d %>% ggplot()+
  geom_line(aes(p_grid, grid.post))+
  geom_col(aes(p_grid, stan.post), alpha = 0.5)+
  labs(x='p_grid', y='posterior')
```







